
---
title: "Infer Spin Vector"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## An example implementation of spin vector component estimation from Statcast data. I'm referencing the following whitepaper by Dr. Alan Nathan

#### https://baseball.physics.illinois.edu/HawkeyeAveSpinComponents.pdf

## Load Necessary Libraries

```{r load-libraries}
library(dplyr)
library(tidyr)
library(readr)
library(plotly)
```

## Calculating Air Density


```{r calculate-air-density}
calculate_air_density <- function(altitude) {
  sea_level_temp <- 288.15 # Standard temperature at sea level in Kelvin
  sea_level_pressure <- 101325 # Standard pressure at sea level in Pascals
  lapse_rate <- 0.0065 # Temperature lapse rate in K/m
  gas_constant <- 287.05 # Specific gas constant for dry air in J/(kg·K)

  temperature <- sea_level_temp - lapse_rate * altitude

  pressure <- sea_level_pressure * (1 - (lapse_rate * altitude / sea_level_temp)) ^ (9.80665 / (gas_constant * lapse_rate))

  air_density <- pressure / (gas_constant * temperature)

  return(air_density)
}
```

## Estimating Drag Coefficient (K)

```{r estimate-drag-coefficient}
estimate_k <- function(altitude = NULL) {
  sea_level_density <- 1.225 # Air density at sea level in kg/m^3
  k0 <- 5.383e-03 # Base drag coefficient

  air_density <- calculate_air_density(altitude)

  k <- k0 * (sea_level_density / air_density)

  return(k)
}
```

## Adding Release Metrics to the Dataset

```{r add-release-metrics}
add_release_metrics <- function(df, altitude_df) {
  z_constant <- 32.174 # Gravity constant in ft/s^2
  
  df <- df %>% 
    left_join(altitude_df, by = 'home_team') # Joining altitude data
  
  df <- df %>%
    mutate(
      yR = 60.5 - release_extension, # Distance to home plate
      tR = (-vy0 - sqrt(vy0^2 - 2 * ay * (50 - yR))) / ay,
      vxR = vx0 + ax * tR,
      vyR = vy0 + ay * tR,
      vzR = vz0 + az * tR,
      tf = (-vyR - sqrt(vyR^2 - 2 * ay * (yR - 17/12))) / ay,
      vxbar = (2 * vxR + ax * tf) / 2,
      vybar = (2 * vyR + ay * tf) / 2,
      vzbar = (2 * vzR + az * tf) / 2,
      vbar = sqrt(vxbar^2 + vybar^2 + vzbar^2),
      adrag = -(ax * vxbar + ay * vybar + (az + z_constant) * vzbar) / vbar,
      amagx = ax + adrag * vxbar / vbar,
      amagy = ay + adrag * vybar / vbar,
      amagz = az + adrag * vzbar / vbar + z_constant,
      amag = sqrt(amagx^2 + amagy^2 + amagz^2),
      K = estimate_k(altitude),
      Cl = amag / (K * vbar^2),
      s = 0.166 * log(0.336 / (0.336 - Cl)),
      spinT = 78.92 * s * vbar,
      spin_efficiency = pmin(spinT / release_spin_rate, 1)
    ) %>%
    drop_na(vx0, vy0, vz0, ax, ay, az, release_extension, release_spin_rate, spin_axis)
  
  return(df)
}
```



## Solving Spin Components

Two methods are implemented to solve for the spin components: `solve_spin_components_method_1` and `solve_spin_components_method_2`. These methods calculate the x, y, and z components of the spin vector based on the given spin efficiency and other parameters.

### Method 1

```{r solve-spin-components-1}
solve_spin_components_method_1 <- function(df){
  w_df <- df %>% 
    mutate(
      phi = spin_axis * pi / 180,
      v = sqrt(vxR^2 + vyR^2 + vzR^2),
      v_hat_x = vxR/v,
      v_hat_y = vyR/v,
      v_hat_z = vzR/v,
      A = v_hat_x * cos(phi) + v_hat_z * sin(phi),
      B = v_hat_y,
      C = spin_efficiency,
      R = sqrt(A^2 + B^2),
      X = atan2(B, A),
      Theta = asin(C / R) - X, 
      spin_x = release_spin_rate * sin(Theta) * cos(phi),
      spin_y = release_spin_rate * cos(Theta),
      spin_z = release_spin_rate * sin(Theta) * sin(phi)
    ) %>% 
    select(spin_x,spin_y,spin_z)
  
  df <- cbind(df,w_df)
  return(df)
}
```

### Method 2

```{r solve-spin-components-2}
solve_spin_components_method_2 <- function(df) {
  w_df <- df %>% 
    mutate(
      phi = spin_axis * pi / 180,
      sign = ifelse(spin_efficiency < 0, 1, -1), # Default is RHP. Flips if LHP
      cos_Theta = sign * abs(1 - spin_efficiency),
      Theta = acos(cos_Theta),
      spin_x = release_spin_rate * sin(Theta) * cos(phi),
      spin_y = release_spin_rate * cos(Theta),
      spin_z = release_spin_rate * sin(Theta) * sin(phi)
    ) %>% 
    select(spin_x,spin_y,spin_z)
  
  df <- cbind(df,w_df)
  return(df)
}
```

## Fetching Data


```{r fetch-data}
library(baseballr)
season_data <- scrape_statcast_savant_pitcher('2024-01-10','2024-08-12',543037)

altitude_data <- read_csv('https://raw.githubusercontent.com/maxbay/public_baseball/main/data/altitude_data.csv')
season_data <- add_release_metrics(season_data,altitude_data)
```

## Applying Spin Component Methods

We now apply the two methods for calculating spin components and compare their results.

```{r apply-spin-methods}
sc_m1 <- solve_spin_components_method_1(season_data)
sc_m2 <- solve_spin_components_method_2(season_data)
```

### Compare Spin Component Methods

```{r method1 try}
sc_m1 %>% 
  dplyr::filter(pitch_type == "FF") %>% 
  select(player_name,pitch_type,spin_x,spin_y,spin_z) %>% 
  head()
```

```{r method2 try}
sc_m2 %>% 
  dplyr::filter(pitch_type == "FF") %>% 
  select(player_name,pitch_type,spin_x,spin_y,spin_z) %>% 
  head()
```

## Averaging Spin Components

This smooths over pitch-level error

```{r average-spin-components}
avg_spin <- sc_m1 %>% 
  group_by(pitcher,player_name,pitch_type) %>% 
  summarise(
    spin_x = mean(spin_x,na.rm = TRUE),
    spin_y = mean(spin_y,na.rm = TRUE),
    spin_z = mean(spin_z,na.rm = TRUE)
  )
```

## Visualizing Spin Components

```{r visualize-spin}
p_type = "FF"
spin_vector <- as.numeric(avg_spin[avg_spin$pitch_type == p_type, c('spin_x', 'spin_y', 'spin_z')])

spin_vector_norm = sqrt(sum(spin_vector^2))

spin_vector <- spin_vector/spin_vector_norm

max_length <- 1  # Adjust to set the range limit

scaled_vector <- spin_vector / max(abs(spin_vector)) * max_length

u <- seq(0, 2*pi, length = 100)
v <- seq(0, pi, length = 100)

x <- outer(cos(u), sin(v))
y <- outer(sin(u), sin(v))
z <- outer(rep(1, length(u)), cos(v))

direction_vector <- c(0, -1, 0) * max_length
  
p <- plot_ly(type = "scatter3d", mode = "lines") %>%
  add_surface(x = x, y = y, z = z, opacity = 0.5, colorscale = list(c(0,1), c("lightblue", "lightblue")), showscale = FALSE) %>%
  add_trace(x = c(-scaled_vector[1], scaled_vector[1]), 
            y = c(-scaled_vector[2], scaled_vector[2]), 
            z = c(-scaled_vector[3], scaled_vector[3]), 
            type = "scatter3d", mode = "lines", line = list(color = 'red', width = 5)) %>%
  add_trace(x = c(0, direction_vector[1]), 
            y = c(0, direction_vector[2]), 
            z = c(0, direction_vector[3]), 
            type = "scatter3d", mode = "lines+markers", line = list(color = 'green', width = 5), 
            marker = list(size = 5, color = 'green', symbol = 'arrow-bar-down-open')) %>%
  layout(scene = list(xaxis = list(title = 'X', range = c(-max_length, max_length)),
                      yaxis = list(title = 'Y', range = c(-max_length, max_length)),
                      zaxis = list(title = 'Z', range = c(-max_length, max_length)),
                      camera = list(eye = list(x = -.8, y = 1.9, z = .8))))

p
```

